---
title: "Final Presentation: Global Oil Trade Networks"
author: "Baxter Lowell"
format: 
  revealjs:
    theme: "sky"
editor: visual
---

```{r}
# read in data (both USA and global oil exports)

library(tidyverse)
library(readr)
library(dplyr)

global_exports_oil <- read.csv(here::here("data/global_oil_exports.csv"), row.names = NULL)

usa_exports <- read.csv(here::here("data/trade_data.csv"), row.names = NULL)

# cleaning up USA oil exports data

usa_exports_clean = usa_exports %>%
  select(c("reporterISO","reporterDesc", "partnerDesc", "partnerISO", "cmdDesc", "altQty", "fobvalue")) %>%
  rename(
    "Exporter" = "reporterISO",
    "Importer" = "partnerISO",
    "Product" = "cmdDesc",
    "quantity_bbl" = "altQty",
    "value_usd" = "fobvalue"
  )%>%
  mutate(
    partnerDesc = recode(partnerDesc, 
                     "Korea" = "Rep. of Korea"
                     )
  )
  

global_exports_clean = global_exports_oil %>%
   select(c("reporterISO", "partnerISO", "reporterDesc", "partnerDesc", "cmdDesc", "altQty", "fobvalue")) %>%
  rename(
    "Exporter" = "reporterISO",
    "Importer" = "partnerISO",
    "Product" = "cmdDesc",
    "quantity_bbl" = "altQty",
    "value_usd" = "fobvalue"
  )%>%
  mutate(
    partnerDesc = recode(partnerDesc, 
                     "Korea" = "Rep. of Korea"
                     )
  )%>%
   slice(-560)
```

## Introduction

For my final project I am looking at global crude oil trade data from 2023. My data comes from the United Nations Comtrade Data set, which is a comprehensive database for global trade, including hundreds of products, imports and exports, and more. I also am using the Happy Planet Index to introduce some additional variables.

## Data/Variables

These are the following variables of interest for my project:

Exporter - The country that is exporting the good.

Importer - The country that is importing the good.

quantity_bbl - Quantity in Barrels of oil.

value_usd - The value of oil traded in USD

Region - Region of the world, assigned to each country.

Life Expectancy, Population, Governance Rank, GDP Per capita.

## Question(s) of Interest

Who does the US export oil to? Does exporting oil prop up an economy well enough to directly relate to a higher quality of life? What do trade networks look like for varying countries across the globe and how can this help explain who is effected most by global events such as conflicts, shortages, etc.?

## Graph 1: USA Crude Oil Exports

```{r}
usa_exports_clean %>%
  filter(Importer != "W00") %>%
  filter(partnerDesc != "Other Asia, nes")%>%
  arrange(desc(quantity_bbl))%>%
  slice_head(n = 20)%>%
  mutate(
    partnerDesc = fct_reorder(partnerDesc, quantity_bbl, .desc = TRUE))%>%

ggplot(data = ., aes(x = partnerDesc, y = quantity_bbl))+
  geom_col()+
  labs(title = "Top 20 US Crude Oil Export Destinations (2023)",
       x = "Partners",
       y = "Annual Barrels (BBL)",
       caption = "*Data from UN Comtrade Database 2023 Data"
       )+
  theme_bw()+
  coord_flip()
```

# Graph 2: Oil Exports vs Life Expectancy

```{r}
library(fuzzyjoin)
library(plotly)
hpi_df <- read_csv("https://raw.githubusercontent.com/highamm/ds234_quarto/main/data_online/hpi-tidy.csv")

global_export_region <- stringdist_full_join(
  hpi_df,
  global_exports_clean,
  by = c("Country" = "reporterDesc"),
  max_dist = 2
)
# graph 2: Relating oil exports to life expectancy

global_export_region1 <- global_export_region %>%
  filter(Importer == "W00")%>%
  filter(quantity_bbl > 0)
  
region_plot <- ggplot(data = global_export_region1, aes(x = LifeExpectancy, y = quantity_bbl, color = Region, label = Country))+
  geom_point()+
  theme_bw()+
  scale_color_viridis_d()+
  labs(title = "Life Expectancy Based on Crude Oil Exports",
       x = " Life Expectancy",
       y = "Barrels Exported",
       caption = "*data from 2023 UN Comtrade Data & Happy Planet Index")

ggplotly(p = region_plot, tooltip = "label")
```

# Graph 3: Network Graph

```{r}
library(rvest)
library(tidygraph)
library(ggraph)


network_graph_data <- global_export_region %>%
  rename(
    from = reporterDesc,
    to = partnerDesc,
    barrels = quantity_bbl
  )

nodes <- network_graph_data %>%
  select(to, from, Country, Region)%>%
  filter(!is.na(from) & !is.na(to)) %>%
  distinct(Country, Region)%>%
  slice(-89)%>%
  rename(name = Country)

valid_nodes <- nodes$name

edges <- left_join(network_graph_data, nodes, by = join_by(Country == name)) %>%
  filter(to != "World") %>%
  select(from, to, barrels) %>%
  filter(!is.na(from) & !is.na(to))

# the following came from Claude, with the prompt : why am I getting an error in my edgelist saying there are NA's present, when I have checked for NA values?

edges <- edges %>%
  filter(from %in% valid_nodes & to %in% valid_nodes)

# End of Claude code

network_obj <- tbl_graph(nodes = nodes, edges = edges, directed = FALSE)

library(ggraph)
ggraph(network_obj |> mutate(centrality = centrality_authority()),
                            layout = "fr") +
  geom_edge_link(aes(width = barrels), alpha = 0.7, show.legend = FALSE) +
  geom_node_point(aes(colour = Region,
                      size = centrality), show.legend = TRUE) +
  scale_edge_width(range = c(0.5, 3)) +
  geom_node_text(aes(label = name), repel = TRUE)+
  scale_colour_viridis_d()
```

Not Done yet: Shiny App of network graph, where user can select a country and an additional variable to color points by.
